Gradual Soundness: Lessons from Static Python

### 2022-01-14 pj-7.1/paper.tex
pj: macro for net CPU change

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\numbenchmark}{2}

@@

  \paragraph{Grounding}
  % What argument, feasibility proof, artifacts,
  % or results and evaluation support this work?
  Our model of \SP{} is implemented in PLT Redex and tested using
  property-based soundness tests and the \SP{} regression suite.
  This paper includes a small core of the model to convey the main ideas
  of the \SP{} approach and its soundness.
  Our performance claims are based on production experience in the Instagram
  web server.
  Thanks to \SP{}, global CPU usage has improved by almost 4\% within a year.

@@

Over the past year, Instagram has been applying gradual soundness to its
primary web server monolith.
The results are very positive.
Although only a handful of modules use \SP{} types (hundreds among thousands),
and only a few critical modules rely on concrete types and primitive types (dozens),
global CPU usage has improved by almost 4\% according to internal profiling tools.
Type-directed optimizations outweigh the cost of enforcing soundness.
Consequently, we believe the details of the \SP{} approach are of interest to
the gradual typing community at large.

@@

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools, which continuously monitor
performance, attribute a 3.7\% improvement in global CPU to \SP{} conversions.
\Cref{t:prod-stat} reports a few other production statistics.
Over 500 modules have been converted thus far.
Despite some initial regressions, none of the converted modules
ran slower after small rewrites and enabling the JIT on certain functions.

\begin{table}
  \caption{Overview: \SP{} in production}
  \label{t:prod-stat}
  \begin{tabular}{l@{~~}l@{\qquad}l@{~~}l}
    %% 541 modules out of ~45000 total
    8     & months in production   & 541   & typed, strict modules \\
    3.7\% & global CPU improvement &   9   & using concrete types \\
    0     & net regressions        &  12   & using primitive types \\
    %% all prims are int32 and int64 and cbool
    %% 873 modules use JIT for some functions (about to increase by OOM)
  \end{tabular}
\end{table}

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

\newcommand{\numbenchmark}{2}
\newcommand{\CPUchange}{3.7\%}

@@

  \paragraph{Grounding}
  % What argument, feasibility proof, artifacts,
  % or results and evaluation support this work?
  Our model of \SP{} is implemented in PLT Redex and tested using
  property-based soundness tests and the \SP{} regression suite.
  This paper includes a small core of the model to convey the main ideas
  of the \SP{} approach and its soundness.
  Our performance claims are based on production experience in the Instagram
  web server.
  Thanks to \SP{}, global CPU usage has improved by \CPUchange{} within a year.

@@

Over the past year, Instagram has been applying gradual soundness to its
primary web server monolith.
The results are very positive.
Although only a handful of modules use \SP{} types (hundreds among thousands),
and only a few critical modules rely on concrete types and primitive types (dozens),
global CPU usage has improved by \CPUchange{} according to internal profiling tools.
Type-directed optimizations outweigh the cost of enforcing soundness.
Consequently, we believe the details of the \SP{} approach are of interest to
the gradual typing community at large.

@@

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools, which continuously monitor
performance, attribute a \CPUchange{} improvement in global CPU to \SP{} conversions.
\Cref{t:prod-stat} reports a few other production statistics.
Over 500 modules have been converted thus far.
Despite some initial regressions, none of the converted modules
ran slower after small rewrites and enabling the JIT on certain functions.

\begin{table}
  \caption{Overview: \SP{} in production}
  \label{t:prod-stat}
  \begin{tabular}{l@{~~}l@{\qquad}l@{~~}l}
    %% 541 modules out of ~45000 total
    8     & months in production   & 541   & typed, strict modules \\
    \CPUchange{} & global CPU improvement &   9   & using concrete types \\
    0     & net regressions        &  12   & using primitive types \\
    %% all prims are int32 and int64 and cbool
    %% 873 modules use JIT for some functions (about to increase by OOM)
  \end{tabular}
\end{table}
